Competing Constructions for German Measure Noun Phrases: from Usage Data to Experimental Validation

### 2017-04-03 Latex/article.tex
new stats, yet againmake all

\subsection{Theoretical assessment}
\label{sec:analyses}

@@

My whole analysis is based on the idea that the syntactic structure [N\Sub{1}~[A~N\Sub{2}]\Sub{NP\Sub{2}}]\Sub{NP\Sub{1}} as instantiated in the \NACa\ and the \PGCa\ is ambiguous.
To show this, the first question is which noun constitutes the head of the whole construction.
This was answered quite clearly by \cite{Loebel1986} already (see also \citealp[213]{Eschenbach1994}, and \citealp[16]{GallmannLindauer1994}).
Subject-verb agreement is always realised on the measure noun, and we can therefore assume that the measure noun is the head.
However, the MNP-internal structure has more interesting cues to offer if the strong\slash weak inflection patterns of adjectives are taken into account.
In NPs with a strongly inflected determiner, attributive adjectives inflect according to the massively syncretistic \textit{weak} pattern.
%\footnote{The weak pattern has only two suffixes.
%The structural cases (nominative and accusative) in the singular are marked with \textit{-e}, the oblique cases (dative and genitive) as well as the entire plural are marked with \textit{-en}.
%The only exception to this rule is the masculine accusative singular, which is also marked with \textit{-en}.}
If there is no determiner (as is the case in the alternating constructions), attributive adjectives inflect like determiners themselves, however.
This is called the \textit{strong} inflectional pattern.
Thus, the adjectives in the \NACa\ and the \PGCa\ have properties of adjectives as well as determiners.
They are lexical adjectives and function as attributive modifiers.
On the other hand, they are inflected like determiners, and they are the leftmost element in the NP, which is typical of determiners.
This unusual double nature of adjectives in NPs without determiners leads to a plausible probabilistic interpretation of the pattern shown in Table~\ref{tab:constructions}.
Whenever speakers classify the adjective in the kind noun phrase more as a determiner, they have to use the \PGCa\ because, if there is a determiner, the PGC (just like in the case of the \PGCd) is the only option.
When they classify the adjective more as an adjective, the kind NP has no determiner, and they have to use the \NACa\ (like in the case of the \NACb).%
\footnote{While the generative analysis presented in \cite{Bhatt1990} cannot properly deal with probabilistic effects, Bhatt comes close to this interpretation by analysing the kind NP in the GPC as a DP and in the NAC as an NP.}

Put differently, the \NACa\ is a \NACb in disguise, and the \PGCa\ is a \PGCd in disguise.
In a probabilistic framework, the crucial question is what controls speakers' decisions to use a a \NACb in disguise or a \PGCd in disguise.
If the \PGCd\ instantiates the PGC prototype and the \NACb\ the NAC prototype, the choice between \PGCa\ and \NACa\ is expected to be influenced by the degree of similarity to either of these prototypes.
I will argue below that the prototypes differ mainly in that they attract different (classes of) measure nouns more or less strongly, mostly because of degrees of grammaticalisation.
However, if lemmas are attracted with different strengths by the two prototypes, then we can expect a baseline attraction effect.
Nouns which occur more often in the \PGCd\ should favour the \PGCa, and nouns which occur more often in the \NACb\ should favour the \NACa.%
\footnote{Such associations between constructions and words or lemmas are the main object of study in approaches such as Collexeme Analysis \citep{GriesStefanowitsch2004}.}
Put differently, a noun's relative frequencies in the prototypical non-alternating constructions should be indicative of same noun's preference to appear in the two alternating variants.
This baseline attraction effect will be quantified in the corpus study Section~\ref{sec:corpusstudies}.

@@

Turning to grammaticalisation, it is often assumed that pseudo-partitives arise as a form of grammaticalised partitives (\eg\ \citealp[536--539]{Koptjevskaja2001} for Finnish and Estonian, \citealp[559]{Koptjevskaja2001} for European languages per se).
%
%\footnote{Notice that the description of German in \citet[547--549]{Koptjevskaja2001} is vague and partly incorrect, maybe because her main source \cite{Eschenbach1993} is not as clear as it could be (see also Footnote \ref{fn:eschenbash} above).
%Her example (29b) from \citet[71]{Eschenbach1993}, repeated here verbatim as (\ref{ex:koptjewskaja}) is misclassified as a clear case of a dative on the measure noun and a genitive on the kind noun (hence, PGC).
%
%\begin{exe}
%  \ex{\label{ex:koptjewskaja}\gll trotz einer Flasche gut-en Wein-s\\
%  in.spite.of one-\textsc{fem.gen} bottle good-\textsc{masc.sg.gen} wine-\textsc{gen}\\
%  \glt in spite of one bottle of good wine}
%\end{exe}
%
%It could, however, equally well be a genitive on the measure noun because of the syncretisms discussed here on p.\ \pageref{page:femininesyncretism} and an ambiguous case requirement of the preposition \textit{troz}, hence an instance of NAC.
%Then, she does not clearly distinguish acceptable from non-acceptable constructions (cf.\ Section \ref{sec:descriptive}) on her pp.\ 548 and 549, including a rather confusing \textit{???} rating of her (31b), which is clearly a non-acceptable construction of the type illustrated in (\ref{ex:analyticalpartitive:b}) above.
%}
%
The alternating constructions in German are both clearly pseudo-partitives, but the grammaticalisation paths uncovered by \cite[esp.\ 526--530]{Koptjevskaja2001} are relevant for an analysis of the alternation.
The grammaticalisation path can start out (in some languages) with constructions involving two referential nouns (not necessarily forming a single and contiguous NP) and a \textit{separative} meaning as in \textit{(cut) two slices from the cake} \citep[535]{Koptjevskaja2001}.
The \textit{part-of} meaning of true partitives as in \textit{a slice of the cake} represents the first stage of a development wherein the measure noun already tends to loose some semantic content.
The pseudo-partitive stage finally instantiates a \textit{quantity-of} relation, potentially even leading to fully grammaticalised quantifiers such as \textit{a lot}.
In German, the PGC is clearly the older variant \citep{Zimmer2015}, still having an overt \textit{quantity-of} marker in the form of the genitive suffix.
Also, it still has the potential to form a true partitive (if the kind noun is definite).
Conversely, the NAC completely lacks this ability to form true partitives, and it forms a more restrictive environment with less combinatory and semantic flexibility.
Hence, we can expect more strongly grammaticalised measure nouns to prefer the NAC.
For example, highly grammaticalised non-referential nouns like \textit{Gramm} `gram' and \textit{Meter} `metre' should occur proportionally more often in the \NACa\ than in the \PGCa.%
\footnote{In case studies like \cite{Brems2003,DeclerckBrems2016} syntactic evidence for grammaticalisation plays a prominent role.
In English -- \cite{Brems2003} argues based on a view from \cite{Langacker1991} -- strongly lexicalised and thus semantically bleached measure nouns turn from nouns into quantifiers, leading to a syntactic reanalysis of the NP, wherein the kind noun becomes the head.
The argument is supported by convincing syntactic evidence, for example subject-verb agreement.
As argued above, in German MNPs the measure noun always remains the syntactic head, and there is no similarly handy criterion to diagnose a high degree of grammaticalisation.}

@@

Furthermore, the grammaticalisation path as described above leads from NPs denoting individuated objects standing in a \textit{part-of} relation to a construction with a more diffuse \textit{quantity-of} relation.
Both types of relations can be numerically quantified -- in as much as a precise number of \textit{parts} or a numerically exact \textit{quantity} can be specified --, but it is much more prototypical of quantities to be specified with numerical precision.
Since the \NACa\ is more closely associated with the \textit{quantity-of} relation, cardinals as attributes of the measure noun are expected to have a higher proportional frequency in the \NACa.
For illustration, (\ref{ex:determiners}) shows the expected variants under this hypothesis.
In (\ref{ex:determiners:nac}), the measure noun is modified by a cardinal \textit{drei} `three', and hence the \NACa\ (kind NP shows case agreement) is preferred.
In (\ref{ex:determiners:pgc}), the measure noun is modified by a non-cardinal determiner \textit{einige} `some', and the \PGCa\ (kind NP is in the genitive) is preferred.

@@

Finally, it has often been iterated that the \PGCa\ is more typical of higher registers or even exclusive to written language (see \citealp[320--323]{Hentschel1993}, where she also refers to descriptive and normative grammars).
This is not surprising in as much as the genitive -- an intrinsic part of the PGC -- is generally underrepresented in colloquial vernacular variants of German as a result of a diachronic process wherein many (but not all) uses of the genitive are replaced by other cases or periphrastic constructions \citep{FleischerSchallert2011}.
Under an integral view of prototypes, which incorporate effects related to larger contexts and registers, such preferences can be to be part of the construction prototypes.
In the corpus study in Section~\ref{sec:corpusstudies}, register effects will therefore be modelled -- even if only using some very simple proxies.

@@

\begin{enumerate}[i.]
  \item The relative frequencies with which measure noun lemmas and kind noun lemmas appear in the prototypical (non-alternating) \PGCd\ and \NACb\ are positively correlated with the relative frequency the alternating \PGCa\ and the \NACa\ are chosen.
  \item (Classes of) more strongly grammaticalised measure nouns favour the NAC.
  \item Measure nouns quantified by cardinals favour the PGC variant.
  \item The NAC is associated more strongly with higher registers.
\end{enumerate}

@@

\begin{table}
  \centering
  \begin{tabular}{llll}
    Unit of reference & Variable                      & Type    & Levels (for factors only)    \\
    \midrule
    Document       & Badness                          & numeric &                              \\
                   & Genitives                        & numeric &                              \\
    Sentence       & \textbf{Construction (response)} & factor  & NACa, PGCa                   \\
                   & Measurecase                      & factor  & Nom, Acc, Dat                \\
                   & Measurenumber                    & factor  & Pl, Sg                       \\
                   & Leftcontext                      & factor  & C, D, O, A, P                \\
    Kind lemma     & Kindattraction                   & numeric &                              \\
                   & Kindfreq                         & numeric &                              \\
    Measure lemma  & Measureattraction                & numeric &                              \\
                   & Measureclass                     & factor  & Physical, Container, Amount, \\
                   &                                  &         & Portion, Rest                \\
                   & Measurefreq                      & numeric &                              \\
  \end{tabular}
  \caption{Annotated variables for the main sample}
  \label{tab:variables}
\end{table}

The variables \textit{Kindattraction} and \textit{Measureattraction} encode the ratio with which a given kind noun lemma or measure noun lemma occurs in the \NACb\ and the \PGCd.
They were calculated from the auxiliary samples described at the end of Section~\ref{sec:gettingdata} as a log-transformed quotient.
The higher the value, the more often the noun occurs in the \PGCa (proportionally).
\textit{Kindfreq} and \textit{Measurefreq} are the logarithm-transformed frequencies per 1,000,000 words of each lemma, extracted from the frequency lists distributed by the DECOW corpus creators on their web page.

@@

The variable \textit{Leftcontext} encodes the coarse-grained part-of-speech of the word preceding the measure noun.
The level names stand for (again with their frequency in the final sample) \textit{cardinal} ($f=1,939$), \textit{determiner} ($f=1,971$), \textit{other} ($f=221$), \textit{adjective} ($f=845$), and \textit{preposition} ($f=87$).
The purpose of this variable is mainly to test whether cardinals really favour the \PGCa\ as hypothesised in Section~\ref{sec:analyses}.
Along the same lines, \textit{Measurenumber} encodes the grammatical case and number of the measure noun.

@@

\begin{figure}[hb!]
  \centering
  \includegraphics[width=0.85\textwidth]{../R/output/corpus_fixeffs_mle+mcmc}
  \caption{Coefficients (MLE and MCMC) with 95\% confidence intervals (for details see text); the intercept (\textit{Leftcontext=C}, \textit{Measurecase=Nom}, \textit{Measurenumber=Sg}, \textit{Kindgender=Masc}, \textit{Measureclass=Physical}; 0 for all numeric z-transformed regressors) is -4.328 (MLE) and -4.441 (MCMC)}
  \label{fig:fixeffs}
\end{figure}

@@

\begin{sidewaystable}
  \centering
  \resizebox{\textheight}{!}{
  \begin{tabular}{llrlp{0.5em}rrp{0.5em}rrp{0.5em}rrp{0.5em}cc}
      Level & Regressor & \multicolumn{1}{l}{\pPB} & Level  && \multicolumn{2}{l}{Coefficient} && \multicolumn{2}{l}{CI low} && \multicolumn{2}{l}{CI high} && \multicolumn{2}{l}{CI contains 0} \\
              &                   &       &           && \multicolumn{1}{l}{MLE} & \multicolumn{1}{l}{MCMC} && \multicolumn{1}{l}{MLE} & \multicolumn{1}{l}{MCMC} && \multicolumn{1}{l}{MLE} & \multicolumn{1}{l}{MCMC} && \multicolumn{1}{l}{MLE} & \multicolumn{1}{l}{MCMC} \\\midrule

    First     & Badness           & 0.001 & (numeric) && -0.165 & -0.169 && -0.264 & -0.264 && -0.074 & -0.075 &&      * &      * \\
              & Genitives         & 0.001 & (numeric) && -0.630 & -0.645 && -0.710 & -0.739 && -0.523 & -0.552 &&      * &      * \\
              & Leftcontext       & 0.001 & D         &&  1.178 &  1.173 &&  0.751 &  0.784 &&  1.548 &  1.546 &&      * &      * \\
              &                   &       & O         &&  1.561 &  1.588 &&  1.093 &  1.179 &&  1.919 &  2.007 &&      * &      * \\
              &                   &       & A         &&  2.300 &  2.340 &&  1.872 &  1.959 &&  2.644 &  2.726 &&      * &      * \\
              &                   &       & P         &&  2.353 &  2.371 &&  1.693 &  1.766 &&  2.900 &  2.978 &&      * &      * \\
              & Measurecase       & 0.001 & Acc       &&  0.065 &  0.070 && -0.130 & -0.119 &&  0.268 &  0.269 &&        &        \\
              &                   &       & Dat       &&  0.707 &  0.735 &&  0.413 &  0.480 &&  0.959 &  1.002 &&      * &      * \\
	      & Measurenumber     & 0.001 & Pl        &&  0.814 &  0.825 &&  0.534 &  0.574 &&  1.080 &  1.079 &&      * &      * \\[0.5\baselineskip]

    Second    & Kindattraction    & 0.033 & (numeric) &&  0.204 &  0.216 &&  0.030 &  0.014 &&  0.366 &  0.412 &&      * &      * \\
    (Kind)    & Kindfreq          & 0.066 & (numeric) &&  0.161 &  0.173 && -0.008 &  0.009 &&  0.333 &  0.350 &&        &      * \\
              & Kindgender        & 0.001 & Neut      && -0.007 & -0.022 && -0.407 & -0.424 &&  0.408 &  0.379 &&        &        \\
	      &                   &       & Fem       &&  1.357 &  1.379 &&  0.831 &  0.872 &&  1.808 &  1.897 &&      * &      * \\[0.5\baselineskip]

    Second    & Measureattraction & 0.015 & (numeric) &&  0.247 &  0.275 &&  0.058 &  0.078 &&  0.413 &  0.476 &&      * &      * \\
    (Measure) & Measureclass      & 0.001 & Rest      &&  0.564 &  0.488 && -0.140 & -0.309 &&  1.197 &  1.236 &&        &        \\
              &                   &       & Container &&  0.609 &  0.618 &&  0.050 &  0.018 &&  1.145 &  1.215 &&      * &      * \\
              &                   &       & Amount    &&  0.927 &  0.959 &&  0.184 &  0.189 &&  1.566 &  1.718 &&      * &      * \\
              &                   &       & Portion   &&  1.472 &  1.503 &&  0.916 &  0.956 &&  1.938 &  2.064 &&      * &      * \\
              & Measurefreq       & 0.003 & (numeric) && -0.257 & -0.264 && -0.413 & -0.435 && -0.098 & -0.099 &&      * &      * \\
  \end{tabular}
  }
  \caption{Coefficient table comparing Maximum Likelihood Estimation (MLE, with 95\% bootstrap confidence interval) and `Bayesian' Markov-Chain Monte Carlo estimation (MCMC); for details see text}
  \label{tab:bigtable}
\end{sidewaystable}

Standard diagnostics for the Maximum Likelihood Estimation (MLE) show that the model quality is quite high.
Generalised variance inflation factors for the regressors were calculated to check for multicollinearity \citep{FoxMonette1992,ZuurEa2010}, and none of the corrected $\text{GVIF}^{1/2\text{df}}$ was higher than $2$.
Nakagawa \& Schielzeth's pseudo-coefficient of determination is $R_m^2=0.423$ and $R^2_c=0.515$ (see \citealp{Gries2015} for a linguist-friendly introduction to these $R^2$ measures, or else \citealp{NakagawaSchielzeth2013}).
The rate of correct predictions is $0.855$, which means a proportional reduction of error of $\lambda=0.354$.
The measure lemma intercepts have a standard deviation of $\sigma_{\text{Measurelemma}}=0.623$, the kind lemma intercept $\sigma_{\text{Kindlemma}}=0.488$.

>>>>

\subsection{Factors controlling the alternation}
\label{sec:analyses}

@@

My whole analysis is based on the idea that the syntactic structure [N\Sub{1}~[A~N\Sub{2}]\Sub{NP\Sub{2}}]\Sub{NP\Sub{1}} as instantiated in the \NACa\ and the \PGCa\ is ambiguous.
To show this, the first question is which noun constitutes the head of the whole construction.
This was answered quite clearly by \cite{Loebel1986} already (see also \citealp[213]{Eschenbach1994}, and \citealp[16]{GallmannLindauer1994}).
Subject-verb agreement is always realised on the measure noun, and we can therefore assume that the measure noun is the head.
However, the MNP-internal structure has more interesting cues to offer if the strong\slash weak inflection patterns of adjectives are taken into account.
In NPs with a strongly inflected determiner, attributive adjectives inflect according to the massively syncretistic \textit{weak} pattern.
%\footnote{The weak pattern has only two suffixes.
%The structural cases (nominative and accusative) in the singular are marked with \textit{-e}, the oblique cases (dative and genitive) as well as the entire plural are marked with \textit{-en}.
%The only exception to this rule is the masculine accusative singular, which is also marked with \textit{-en}.}
If there is no determiner (as is the case in the alternating constructions), attributive adjectives inflect like determiners themselves, however.
This is called the \textit{strong} inflectional pattern.
Thus, the adjectives in the \NACa\ and the \PGCa\ have properties of adjectives as well as determiners.
On the one hand, they are lexical adjectives and function as attributive modifiers.
On the other hand, they are inflected like determiners, and they are the leftmost element in the NP, which is typical of determiners.
This unusual double nature of adjectives in NPs without determiners leads to a plausible probabilistic interpretation of the pattern shown in Table~\ref{tab:constructions}.
Whenever speakers classify the adjective in the kind noun phrase more as a determiner, they have to use the \PGCa\ because, if there is a determiner, the PGC (just like in the case of the \PGCd) is the only option.
When they classify the adjective more as an adjective, the kind NP has no determiner, and they have to use the \NACa\ (like in the case of the \NACb).%
\footnote{While the generative analysis presented in \cite{Bhatt1990} cannot properly deal with probabilistic effects, Bhatt comes close to this interpretation by analysing the kind NP in the GPC as a DP and in the NAC as an NP.}

This morpho-syntactic ambiguity means that the \NACa\ is in fact a \NACb in disguise, and the \PGCa\ is a \PGCd in disguise.
This should ideally be reflected in the following basic selection effect:
It is known from frameworks like Collexeme Analysis \citep{GriesStefanowitsch2004} that lemmas are attracted with different strengths by competing constructions.
If the \NACa\ is highly similar to the \NACb\ and the \PGCa\ is highly similar to the \PGCd, the probability with which individual noun lemmas appear in the alternating constructions should be predictable at least partly from the relative frequency with which the same lemmas are used in the non-alternating constructions.
Nouns which occur proportionally more often in the \PGCd\ should favour the \PGCa, and nouns which occur proportionally more often in the \NACb\ should favour the \NACa.
This basic attraction effect will be quantified in the corpus study Section~\ref{sec:corpusstudies}.

@@

In a probabilistic framework like Prototype Theory, the crucial question is, however, what controls speakers' decisions to use either variant beyond simple lemma preference effects. 
I argue that the NAC and the PGC are prototypes associated with different degrees of the grammaticalisation of the measure noun, morpho-syntactic properties, and register effects.
The degree of similarity of a given instance to either of the two prototypes makes speakers chose the \NACa\ or the \PGCa.
I first turn to grammaticalisation.
It is often assumed that pseudo-partitives arise as a form of grammaticalised partitives (\eg\ \citealp[536--539]{Koptjevskaja2001} for Finnish and Estonian, \citealp[559]{Koptjevskaja2001} for European languages per se).
%
%\footnote{Notice that the description of German in \citet[547--549]{Koptjevskaja2001} is vague and partly incorrect, maybe because her main source \cite{Eschenbach1993} is not as clear as it could be (see also Footnote \ref{fn:eschenbash} above).
%Her example (29b) from \citet[71]{Eschenbach1993}, repeated here verbatim as (\ref{ex:koptjewskaja}) is misclassified as a clear case of a dative on the measure noun and a genitive on the kind noun (hence, PGC).
%
%\begin{exe}
%  \ex{\label{ex:koptjewskaja}\gll trotz einer Flasche gut-en Wein-s\\
%  in.spite.of one-\textsc{fem.gen} bottle good-\textsc{masc.sg.gen} wine-\textsc{gen}\\
%  \glt in spite of one bottle of good wine}
%\end{exe}
%
%It could, however, equally well be a genitive on the measure noun because of the syncretisms discussed here on p.\ \pageref{page:femininesyncretism} and an ambiguous case requirement of the preposition \textit{troz}, hence an instance of NAC.
%Then, she does not clearly distinguish acceptable from non-acceptable constructions (cf.\ Section \ref{sec:descriptive}) on her pp.\ 548 and 549, including a rather confusing \textit{???} rating of her (31b), which is clearly a non-acceptable construction of the type illustrated in (\ref{ex:analyticalpartitive:b}) above.
%}
%
The alternating constructions in German are both clearly pseudo-partitives, but the grammaticalisation paths uncovered by \cite[esp.\ 526--530]{Koptjevskaja2001} are still relevant in the case at hand.
The grammaticalisation path can start out (in some languages) with constructions involving two referential nouns (not necessarily forming a single and contiguous NP) and a \textit{separative} meaning as in \textit{(cut) two slices from the cake} \citep[535]{Koptjevskaja2001}.
The \textit{part-of} meaning of true partitives as in \textit{a slice of the cake} represents the first stage of a development wherein the measure noun already tends to loose some semantic content.
The pseudo-partitive stage finally instantiates a \textit{quantity-of} relation, potentially even leading to fully grammaticalised quantifiers such as \textit{a lot}.
In German, the PGC is clearly the older variant \citep{Zimmer2015}.
It still has the potential to form a true partitive (if the kind noun is definite).
Conversely, the NAC completely lacks this ability to form true partitives, and it forms a more restrictive environment with less combinatory and semantic flexibility.
Hence, we can expect the \NACa\ to be a prototypical hosting construction for more strongly grammaticalised measure nouns.
For example, highly grammaticalised non-referential nouns like \textit{Gramm} `gram' and \textit{Meter} `metre' should occur proportionally more often in the \NACa\ than in the \PGCa.
% \footnote{In case studies like \cite{Brems2003,DeclerckBrems2016} syntactic evidence for grammaticalisation plays a prominent role.
% In English -- \cite{Brems2003} argues based on a view from \cite{Langacker1991} -- strongly lexicalised and thus semantically bleached measure nouns turn from nouns into quantifiers, leading to a syntactic reanalysis of the NP, wherein the kind noun becomes the head.
% The argument is supported by convincing syntactic evidence, for example subject-verb agreement.
% As argued above, in German MNPs the measure noun always remains the syntactic head, and there is no similarly handy criterion to diagnose a high degree of grammaticalisation.}

@@

Furthermore, the grammaticalisation path as described above leads from NPs denoting individuated objects standing in a \textit{part-of} relation to a construction with a more diffuse \textit{quantity-of} relation.
Both types of relations can be numerically quantified -- in as much as a precise number of \textit{parts} or a numerically exact \textit{quantity} can be specified --, but it is much more prototypical of quantities to be specified with numerical precision.
Since the \NACa\ is more closely associated with the \textit{quantity-of} relation, cardinals as attributes of the measure noun are expected to have a higher proportional frequency in the \NACa.
For illustration, (\ref{ex:determiners}) shows the expected variants under this hypothesis.
In (\ref{ex:determiners:nac}), the measure noun is modified by a cardinal \textit{drei} `three', and hence the \NACa\ (the kind NP shows case agreement) is preferred.
In (\ref{ex:determiners:pgc}), the measure noun is modified by a non-cardinal determiner \textit{einige} `some', and the \PGCa\ (the kind NP is in the genitive) is preferred.

@@

Finally, it has often been iterated that the \PGCa\ is more typical of higher registers or even exclusive to written language (see \citealp[320--323]{Hentschel1993}).
This is not surprising in as much as the genitive -- an intrinsic part of the PGC -- is generally underrepresented in colloquial vernacular variants of German as a result of a diachronic process wherein many (but not all) uses of the genitive are replaced by other cases or periphrastic constructions \citep{FleischerSchallert2011}.
Under an integral view of prototypes, which incorporate effects related to larger contexts and registers, such preferences can be to be part of the construction prototypes.
In the corpus study in Section~\ref{sec:corpusstudies}, register effects will therefore be modelled -- even if only using some very simple proxies.

@@

\begin{enumerate}[i.]
  \item The relative frequencies with which measure noun lemmas and kind noun lemmas appear in the prototypical (non-alternating) \PGCd\ and \NACb\ are predictive of the probability with which the alternating \PGCa\ and the \NACa\ are chosen.
  \item (Classes of) more strongly grammaticalised measure nouns favour the \NACa.
  \item Measure nouns modified by cardinals favour the \NACa.
  \item The \NACa\ is associated with higher registers.
\end{enumerate}

@@

\begin{table}
  \centering
  \begin{tabular}{llll}
    Unit of reference & Variable                      & Type    & Levels (for factors only) \\
    \midrule
    Document       & Badness                          & numeric &                           \\
                   & Genitives                        & numeric &                           \\
    Sentence       & Cardinal                         & factor  & Yes, No                   \\
                   & \textbf{Construction (response)} & factor  & NACa, PGCa                \\
                   & Measurecase                      & factor  & Nom, Acc, Dat             \\
    Kind lemma     & Kindattraction                   & numeric &                           \\
                   & Kindfreq                         & numeric &                           \\
    Measure lemma  & Measureattraction                & numeric &                           \\
                   & Measureclass                     & factor  & Physical, Container,      \\
                   &                                  &         & Amount, Portion, Rest     \\
                   & Measurefreq                      & numeric &                           \\
  \end{tabular}
  \caption{Annotated variables for the main sample}
  \label{tab:variables}
\end{table}

The variables \textit{Kindattraction} and \textit{Measureattraction} encode the ratio with which a given kind noun lemma or measure noun lemma occurs in the \NACb\ and the \PGCd.
They were calculated from the auxiliary samples described at the end of Section~\ref{sec:gettingdata} as a log-transformed quotient.
The higher the value, the more often the noun occurs in the \PGCa (proportionally).%
\footnote{One could argue that some more advanced measure of attraction strength should be used, as is done in Collostructional (or Collexeme) Analysis \citep{GriesStefanowitsch2004}.
Two main points speak against such an approach.
First, the attraction values will be used as regressors in a GLMM, and the values resulting from collostructional approaches, \ie\ logarithmised Fisher p values, have a very unfavourable distribution for such a use.
They cluster around $-\infty$ and $0$, especially when (as in this case) cells contain $0$ count values.
This cannot be remedied on principled grounds, since, for example, a z transformation is inadequate for such a strongly bimodal distribution (even if we clamp $-\infty$ to a very low numeric value).
Second, their main use in collostructional analyses is to \textit{sort} a list of collexemes and interpret the order.
Their concrete numerical value thus does not have a solid cognitive interpretation.
For the present study, it suffices to estimate the relative frequency with which speakers have encountered a specific lemma in the \NACa\ and the \PGCa.
}
\textit{Kindfreq} and \textit{Measurefreq} are the logarithm-transformed frequencies per 1,000,000 words of each lemma, extracted from the frequency lists distributed by the DECOW corpus creators on their web page.

@@

The variable \textit{Cardinal} encodes whether the measure noun is modified by a cardinal ($f=1,939$) or not ($f=3,124$).
The purpose of this variable is to test whether cardinals really favour the \NACa\ as hypothesised in Section~\ref{sec:analyses}.

@@

\begin{figure}[hb!]
  \centering
  \includegraphics[width=0.85\textwidth]{../R/output/corpus_fixeffs_mle+mcmc}
  \caption{Coefficients (MLE and MCMC) with 95\% confidence intervals (for details see text); the intercept (\textit{Cardinal=Yes}, \textit{Measurecase=Nom}, \textit{Kindgender=Masc}, \textit{Measureclass=Physical}; 0 for all numeric z-transformed regressors) is -4.328 (MLE) and -4.441 (MCMC)}
  \label{fig:fixeffs}
\end{figure}

@@

\begin{sidewaystable}
  \centering
  \resizebox{\textheight}{!}{
  \begin{tabular}{llrlp{0.5em}rrp{0.5em}rrp{0.5em}rrp{0.5em}cc}
      Level & Regressor & \multicolumn{1}{l}{\pPB} & Level  && \multicolumn{2}{l}{Coefficient} && \multicolumn{2}{l}{CI low} && \multicolumn{2}{l}{CI high} && \multicolumn{2}{l}{CI contains 0} \\
              &                   &       &           && \multicolumn{1}{l}{MLE} & \multicolumn{1}{l}{MCMC} && \multicolumn{1}{l}{MLE} & \multicolumn{1}{l}{MCMC} && \multicolumn{1}{l}{MLE} & \multicolumn{1}{l}{MCMC} && \multicolumn{1}{l}{MLE} & \multicolumn{1}{l}{MCMC} \\\midrule
       First     & Badness           &  0.002 &           && -0.152 & -0.155 && -0.247 & -0.247 && -0.061 & -0.065 && * & * \\
                 & Cardinal          &  0.001 & No        &&  1.189 &  1.222 &&  0.862 &  0.927 &&  1.466 &  1.496 && * & * \\
                 & Genitives         &  0.001 &           && -0.693 & -0.711 && -0.768 & -0.801 && -0.592 & -0.616 && * & * \\
                 & Measurecase       &  0.001 & Acc       &&  0.030 &  0.031 && -0.150 & -0.159 &&  0.212 &  0.222 &&   &   \\
                 &                   &        & Dat       &&  0.705 &  0.729 &&  0.455 &  0.465 &&  0.944 &  0.995 && * & * \\[0.5\baselineskip]

       Second    & Kindattraction    &  0.020 &           &&  0.225 &  0.244 &&  0.049 &  0.056 &&  0.393 &  0.422 && * & * \\
       (Kind)    & Kindfreq          &  0.095 &           &&  0.146 &  0.164 && -0.023 & -0.016 &&  0.301 &  0.341 &&   &   \\
                 & Kindgender        &  0.001 & Neut      &&  0.021 &  0.013 && -0.367 & -0.409 &&  0.392 &  0.435 &&   &   \\
                 &                   &        & Fem       &&  1.269 &  1.289 &&  0.800 &  0.788 &&  1.709 &  1.783 && * & * \\[0.5\baselineskip]

       Second    & Measureattraction &  0.001 &           &&  0.282 &  0.299 &&  0.106 &  0.102 &&  0.447 &  0.515 && * & * \\
       (Measure) & Measureclass      &  0.001 & Container &&  0.252 &  0.257 && -0.265 & -0.303 &&  0.788 &  0.813 &&   &   \\
                 &                   &        & Rest      &&  0.421 &  0.379 && -0.209 & -0.378 &&  1.063 &  1.091 &&   &   \\
                 &                   &        & Amount    &&  0.831 &  0.889 &&  0.215 &  0.220 &&  1.432 &  1.569 && * & * \\
                 &                   &        & Portion   &&  1.217 &  1.253 &&  0.675 &  0.689 &&  1.684 &  1.840 && * & * \\
                 & Measurefreq       &  0.005 &           && -0.231 & -0.232 && -0.363 & -0.395 && -0.079 & -0.073 && * & * \\

  \end{tabular}
  }
  \caption{Coefficient table comparing Maximum Likelihood Estimation (MLE, with 95\% bootstrap confidence interval) and `Bayesian' Markov-Chain Monte Carlo estimation (MCMC); the intercept (\textit{Cardinal=Yes}, \textit{Measurecase=Nom}, \textit{Kindgender=Masc}, \textit{Measureclass=Physical}; 0 for all numeric z-transformed regressors) is -3.548 (MLE) and -3.700 (MCMC)}
  \label{tab:bigtable}
\end{sidewaystable}

Standard diagnostics for the Maximum Likelihood Estimation (MLE) show that the model quality is quite high.
Generalised variance inflation factors for the regressors were calculated to check for multicollinearity \citep{FoxMonette1992,ZuurEa2010}, and none of the corrected $\text{GVIF}^{1/2\text{df}}$ was higher than $1.6$.
Nakagawa \& Schielzeth's pseudo-coefficient of determination is $R_m^2=0.409$ and $R^2_c=0.495$ (see \citealp{Gries2015} for a linguist-friendly introduction to these $R^2$ measures, or else \citealp{NakagawaSchielzeth2013}).
The rate of correct predictions is $0.843$, which means a proportional reduction of error of $\lambda=0.297$.
The measure lemma intercepts have a standard deviation of $\sigma_{\text{Measurelemma}}=0.448$, the kind lemma intercept $\sigma_{\text{Kindlemma}}=0.604$.

The coefficient estimates are specified in Table~\ref{tab:bigtable} for each regressor (or regressor level) in the columns labelled \textit{Coefficient}.
For a robust quantification of the precision of the estimation, I ran a parametric bootstrap (using the \textit{confint.merMod} function from \textit{lme4}) with $1,000$ replications, and using the percentile method for the calculation of the intervals.
The resulting 95\% bootstrap confidence intervals are reported in Table~\ref{tab:bigtable} in the columns labelled \textit{CI low} and \textit{CI high} (= upper and lower 2.5th percentiles).
The column \textit{CI contains 0} shows an asterisk for those intervals that do \textit{not} include 0.
Furthermore, for each regressor, a $p$ value was obtained by dropping the regressor from the full model, re-estimating the nested model and comparing it to the full model.
Instead of inexact Wald approximations or Likelihood Ratio Tests, I used a drop-in bootstrap replacement for the Likelihood Ratio Test from the function \textit{PBmodcomp} from the \textit{pbkrtest} package \citep{HalekohHojsgaard2014}, hence I call the value $p_{\text{PB}}$.
These bootstrapped $p$ values are given in the columns labelled p\Sub{PB} in Table~\ref{tab:bigtable}.
Only \textit{Kindfreq} ($\mpPB=0.095$) can be seen as slightly too high to be convincing (`non-significant').

Finally, I show that Bayesian methods in the form of Markov-Chain Monte Carlo estimation do not necessarily (and in this case predictably) lead to different results.
The model was re-estimated using the \textit{stan\_glmer} function from the \textit{rstanarm} package, which provides an \textit{lme4}-compatible syntax for estimating common model types with the \textit{stan} software \citep{CarpenterEa2017}.
The algorithm was run with $4$ chains and $1,000$ iterations, and instead of guessing prior distributions or using implausible uniform prior distributions \cite[251--252]{Levshina2016}, I used plausible default priors.
Most notably, priors for coefficients were specified as $\mathcal{N}(0,10)$ because coefficients higher than $10$ or lower than $-10$ are extremely rare in well-specified models on appropriate data.%
\footnote{Consider that with a coefficient of $10$, each increase by $1$ in the regressor variable means an increase in odds of $exp(10)=22,026.47$.
To reliably estimate such coefficients, extremely large samples would be required.}
The algorithm converged, and for all coefficients, the $\hat{R}$ diagnostic was exactly $1$.
The resulting coefficients and intervals as well as an * are also given in Table~\ref{tab:bigtable} in the columns labelled \textit{MCMC}.
Both methods lead to exactly the same results (minus negligible numerical differences) as expected (see Section~\ref{sec:rightstatistics}).
The signs and magnitudes of the coefficients are identical, and confidence intervals have the same width and symmetry properties.
Figure~\ref{fig:fixeffs} illustrates this by also showing both estimates.
Since both estimators converge, I only interpret the MLE model in the next section.
